---
title: "Segmented signal vs whole signal"
author: "Gely Ancelin"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
    output_file: "gely.ancelin.html"
---

<div style="text-align: justify;">

 
The scientific context and the protocol are described in the readme file of the GitHub repository. The SPARC and NNP for the whole signal and segmented signal were computed in the Python project, and the results are presented in the Excel file "results_all.xlsx". 


# Rstudio project  

The aim of this code is to load the SPARC and NNP for the whole signal and segmented signal computed in the Python_project and look for a relationship. 

## Data organisation 


### File name 


**results_repetition.xlsx**
These data are the results of the Python_project, in which we computed the SPARC and NNP for the whole signal and segmented signal for 10 individuals with low back pain performing a flexion-extension task. 

### File structure


The data are organized in an Excel file with the following columns:

- Signal name : FlexionDos_00X_av_repY, where X is the participant number (from 1 to 10) and Y is the repetition number (from 1 to 5) 

- SPARC_flexion : SPARC value calculated for the segmented signal (for each repetition)

- SPARC_extension : SPARC value calculated for the segmented signal (for each repetition)

- NNP_flexion : NNP value calculated for the segmented signal (for each repetition)

- NNP_extension : NNP value calculated for the segmented signal (for each repetition)

- SPARC : SPARC value calculated for the whole signal (for each participant)

- NNP : NNP value calculated for the whole signal (for each participant)

# Notebook organisation



In this notebook, we will perform the following steps : 

1. Import libraries 

2. Load and visualize the data computed in the Python project

3. Create a new dataframe with the mean SPARC and NNP for the segmented signal for each participant

4. Plot the relationship between mean_NNP_flexion/NNP and mean_SPARC_flexion/SPARC

5. Compute the Bland-Altman test for NNP and SPARC to evaluate the agreement between the whole signal and the segmented signal for both metrics.



# Import libraries

```{r}
# import libraries to read excel and compute a plot
library(readxl)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(stringr)

```

# Load and visualise the data computed in the Python project

```{r}
# From data, import "results_repetition.xlsx"
data_subject <- read_excel("data/results_all.xlsx")
data_subject
```






# Visualisation - Segmented signal vs whole signal

## plot NNP as a function of NNP_flexion

This plot shows the relationship between NNP_flexion and NNP. 

- The points represent the data for each participant and each repetition. 

- The red line represents the loess fit, which is a non-parametric method that fits a smooth curve to the data.

- The blue line represents the linear regression fit. 

- The equation of the linear regression and the R-squared value are also displayed on the plot.

```{r}
# Plot NNP_flexion as a function of NNP
ggplot(data_subject, aes(x = mean_NNP_flexion, y = NNP)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  geom_abline(slope = 1, intercept = 0, color = "green", linetype = "dashed") +
  # add the equation and the R-squared value to the plot
  stat_regline_equation(label.x = 3, label.y = 0.9) +
  stat_cor(method = "pearson", label.x = 3, label.y = 0.5) +
  labs(title = "NNP_flexion vs NNP",
       x = "mean_NNP_flexion",
       y = "NNP") +
  theme_minimal()
```



**Observation** : There is a positive correlation between NNP_flexion and NNP. Le loess fit (red line) and the linear fit (blue line) are quite similar, indicating that the relationship between NNP_flexion and NNP is approximately linear.

**Results**: 
the slope value (0.97) indicates that for every unit increase the segmented signal, there is an expected increase of 0.97 units in the whole signal. The intercept value (0.36) suggests that when NNP_flexion is zero, NNP is expected to be 0.36. 

The R-squared value (0.96) indicates that approximately 96% of the variability in NNP_flexion can be explained by the linear relationship with NNP. This suggests that the number of peaks in the segmented signal (NNP_flexion) is strongly correlated with the number of peaks in the whole signal (NNP).

**Interpretation**
This suggests that the whole signal and the segmented signal provide similar information about movement smoothness when using the NNP metric. 

The non-zero intercept value (0.36) might be due to the fact that the whole signal includes resting period in which postural adjustments can occur, which can increase the number of peaks in the whole signal compared to the segmented signal.

## plot SPARC_flexion as a function of SPARC

```{r}
# Plot SPARC_flexion as a function of SPARC
ggplot(data_subject, aes(x = mean_SPARC_flexion, y = SPARC)) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  #geom_abline(slope = 1, intercept = 0, color = "green", linetype = "dashed") +
  stat_regline_equation(label.x = -9, label.y = -13) +
  stat_cor(method = "pearson", label.x = -9, label.y = - 14) +  
  labs(title = "SPARC_flexion vs SPARC",
       x = "mean_SPARC_flexion",
       y = "SPARC") +
  theme_minimal()
  
```

**Observation** : No clear linear relationship between SPARC_flexion and SPARC. The loess fit (red line) shows a non-linear pattern. 

**Results** : The slope value (0.37) indicates that for every unit increase in SPARC_flexion, there is an expected increase of 0.37 units in SPARC. The intercept value (-7.1) suggests that when SPARC_flexion is zero, SPARC is expected to be -7.1.

**Interpretation** : This suggests that the whole signal and the segmented signal provide different information about movement smoothness when using the SPARC metric. 

This difference might be due to the fact that SPARC is more sensitive to movement duration and structure than NNP, which can lead to different values for the whole signal and the segmented signal. 


# Bland-Altman test 

Correlation quantifies the degree to which two variables are linearly related, but it does not provide information about the agreement between the two methods. The bland-Altman test is a method to evaluate the agreement between two methods of measurement (7). 

It consists in plotting the difference between the two methods against the mean of the two methods, and calculating the bias (mean difference) and the limits of agreement (bias ± 1.96 * standard deviation of the differences). 

The difference has to be normally distributed, that's the reason whya shapiro wilk test is performed (7). 

$$difference = whole - segmented$$
Result of the Bland-Altman test will be presented in a table with the following columns :

- bias : mean difference between the whole signal and the segmented signal. This value indicates if there is a systematic difference between the two methods. A bias close to zero indicates good agreement between the two methods, while a bias different from zero indicates a systematic difference.


$$bias = mean(difference)$$
- Upper : upper limit of agreement (bias + 1.96 * sd). This value indicates the upper bound of the range in which most of the differences between the two methods are expected to lie. If the upper limit is close to zero, it suggests that there is good agreement between the two methods for high values.

$$\text{upper limit of agreement} = bias + 1.96 * sd$$
- lower : lower limit of agreement (bias - 1.96 * sd). This value indicates the lower bound of the range in which most of the differences between the two methods are expected to lie. If the lower limit is close to zero, it suggests that there is good agreement between the two methods for low values.

$$\text{lower limit of agreement} = bias - 1.96 * sd$$

- perc_bias : percentage bias (bias / mean of the two methods) * 100. This value indicates the bias as a percentage of the mean of the two methods. A smaller percentage bias indicates better agreement between the two methods. 

$$\text{percentage bias} = \left( \frac{bias}{\text{mean of the two methods}} \right) * 100$$

The result is a graph in which the x-axis represents the mean of the two methods and the y-axis represents the difference between the two methods. The bias is represented by a horizontal line, and the limits of agreement are represented by two dashed lines. 95% of the points should lie within the limits of agreement if the two methods are in good agreement (7)

The interpretation of this graph relies on the line of equality that should be within the limits of agreement, and on the limit of acceptable difference that should be defined a priori based on relevant criteria. (7)

## Bland-Altman - NNP

```{r}
# Compute the Bland-Altman test

data_subject$mean_NNP <- (data_subject$NNP + data_subject$mean_NNP_flexion) / 2
data_subject$diff_NNP <- data_subject$NNP - data_subject$mean_NNP_flexion

# Shapiro wilk test to test the normal distribution of the differences 
shapiro.test(data_subject$diff_NNP)

bias_NNP <- mean(data_subject$diff_NNP, na.rm = TRUE)
sd_NNP <- sd(data_subject$diff_NNP, na.rm = TRUE)

upper_NNP <- bias_NNP + 1.96 * sd_NNP
lower_NNP <- bias_NNP - 1.96 * sd_NNP

# Plot
ggplot(data_subject, aes(x = mean_NNP, y = diff_NNP)) +
  geom_point(linewidth = 3) +
  geom_hline(yintercept = bias_NNP, color = "blue", size = 1) +
  geom_hline(yintercept = upper_NNP, linetype = "dashed", color = "red") +
  geom_hline(yintercept = lower_NNP, linetype = "dashed", color = "red") +
  labs(title = "Bland-Altman Plot (NNP)",
       x = "Mean of methods",
       y = "Difference (Global - Segmented)") +
  theme_minimal()

```


## Bland-Altman test - SPARC

```{r}

# Compute the Bland-Altman test
data_subject$mean_SPARC <- (data_subject$SPARC + data_subject$mean_SPARC_flexion) / 2
data_subject$diff_SPARC <- data_subject$SPARC - data_subject$mean_SPARC_flexion

# Shapiro wilk test to test the normal distribution of the differences 
shapiro.test(data_subject$diff_SPARC)

bias_SPARC <- mean(data_subject$diff_SPARC, na.rm = TRUE)
sd_SPARC <- sd(data_subject$diff_SPARC, na.rm = TRUE)

upper_SPARC <- bias_SPARC + 1.96 * sd_SPARC
lower_SPARC <- bias_SPARC - 1.96 * sd_SPARC

# Plot
ggplot(data_subject, aes(x = mean_SPARC, y = diff_SPARC)) +
  geom_point(linewidth = 3) +
  geom_hline(yintercept = bias_SPARC, color = "blue", size = 1) +
  geom_hline(yintercept = upper_SPARC, linetype = "dashed", color = "red") +
  geom_hline(yintercept = lower_SPARC, linetype = "dashed", color = "red") +
  labs(title = "Bland-Altman Plot (SPARC)",
       x = "Mean of methods",
       y = "Difference (Global - Segmented)") +
  theme_minimal()

```


```{r}
# % bias NNP
mean_NNP_all <- mean(c(data_subject$NNP, data_subject$mean_NNP_flexion), na.rm = TRUE)
perc_bias_NNP <- (bias_NNP / mean_NNP_all) * 100

# % bias SPARC
mean_SPARC_all <- mean(c(data_subject$SPARC, data_subject$mean_SPARC_flexion), na.rm = TRUE)
perc_bias_SPARC <- (bias_SPARC / mean_SPARC_all) * 100

ba_table <- data.frame(
  methode = c("NNP", "SPARC"),
  bias = c(bias_NNP, bias_SPARC),
  sd = c(sd_NNP, sd_SPARC),
  lower = c(lower_NNP, lower_SPARC),
  upper = c(upper_NNP, upper_SPARC), 
  perc_bias = c(perc_bias_NNP, perc_bias_SPARC) 
  )

ba_table
```
The best way to interpret these results would be to define a priori the limit of maximal acceptable difference, based on relevant criteria. Unfortunately, as our method is rarely used, mostly in a flexion movement, there is no consensus about this limit. However, we can look at the percentage bias to have an idea of the agreement between the two methods.

The percentage bias is a measure of the systematic difference between the two methods relative to the mean of the two methods. In that case, the percentage bias is higher for SPARC (49,4%) than for NNP (-5,95%). It means that the NNP for the whole signal is underestimating the NNP for the segmented signal by 5,95%, while the SPARC for the whole signal is overestimating the SPARC for the segmented signal by 49,4%. 

This result is consistent with the correlation results and support our hypothesis. 

# Discussion :

The weak correlation between SPARC_flexion and SPARC, as well as the high percentage bias, suggest that the whole signal and the segmented signal give different values. It confirms the fact that SPARC should not be computed on the whole signal (4). 

The bias for NNP is moderate (-5,95%). It might be due to the fact that the whole signal includes resting periods that may decrease the number of peaks in the whole signal compared to the segmented signal, which only includes the flexion movement. Moreover, the whole signal contains concentric and eccentric movement which lie on different motor control strategies and can have different NNP values. It would be interesting to look at the NNP for the extension movement to see if it is different from the NNP for the flexion movement, and if it can explain the bias between the whole signal and the segmented signal. Finally, it would be interesting to perform a sensitivity analysis to see if prominence chosen in the NNP calculation can influence the bias between the whole signal and the segmented signal.

However, the bias is relatively low and might be acceptable, depending on the context and the criteria for maxiaml acceptable difference.


# Perspective : 

Future works should focus on determine a limit of maximal acceptable difference for both NNP and SPARC. This would allow us to better interpret the results of the Bland-Altman test and to determine if the agreement between the whole signal and the segmented signal is acceptable for clinical use.

# Conclusion

the NNP metric shows a strong correlation between the whole signal and the segmented signal, with a low percentage bias (-5,95%).The SPARC metric shows a weaker correlation between the whole signal and the segmented signal, with a higher percentage bias (49,4%). 

Concerning our study on **the evolution of movement smoothness in trunck flexion before and after a rehabilitation program in individuals with low back pain** : We confirm that the SPARC metric should not be computed on a whole signal. Some caution should be taken when using the whole signal as a proxy for the segmented signal for NNP metric, as there is a moderate bias between the two methods. However, the strong correlation and the relatively low percentage bias suggest that the whole signal might be an acceptable proxy for the segmented signal when using the NNP metric, depending on the context and the criteria for maximal acceptable difference.

# References : 

1. Hartvigsen J, Hancock MJ, Kongsted A, Louw Q, Ferreira ML, Genevay S, et al. What low back pain is and why we need to pay attention. Lancet Lond Engl. 9 juin 2018;391(10137):2356‑67

2. Errabity A, Calmels P, Han WS, Bonnaire R, Pannetier R, Convert R, et al. The effect of low back pain on spine kinematics: A systematic review and meta-analysis. Clin Biomech. août 2023;108:106070

3. Balasubramanian S, Melendez-Calderon A, Roby-Brami A, Burdet E. On the analysis of movement smoothness. J Neuroengineering Rehabil. 9 déc 2015;12:112

4. Balasubramanian S, Melendez-Calderon A, Roby-Brami A, Burdet E. On the analysis of movement smoothness. J Neuroeng Rehabil. 9 déc 2015;12:112. doi:10.1186/s12984-015-0090-9 PubMed PMID: 26651329; PubMed Central PMCID: PMC4674971.

5. Melendez-Calderon A, Shirota C, Balasubramanian S. Estimating Movement Smoothness From Inertial Measurement Units. Front Bioeng Biotechnol. 14 janv 2021;8:558771. doi:10.3389/fbioe.2020.558771 

6. https://github.com/ancelingely/Projet_R_Python_M2_REHAB.git

7. Giavarina D. Understanding Bland Altman analysis. Biochem Med. 2015;25(2):141‑51. doi:10.11613/BM.2015.015





</div>



